---
title: "Modelling Tree Canopy Height using Landsat 8"
format: 
  html: 
   theme: journal
   embed-resources: true
bibliography: ["refs.bibtex", "packages.bib"]
nocite: |
  @*
---
#### (By Hugh Graham)

## Introduction

This workbook outlines a minimal example, using the R programming language,
for generating a canopy height model from Earth Observation (EO) data using 
Machine Learning. This workbook should be reproducible providing that the 
required packages are installed. 

The workbook will demonstrate how to download training data from a Zenodo 
repository, generate a coincident Landsat 8 annual composite map and apply 
a short machine learning pipeline done locally in R with no need for cloud 
compute. However, the scale of the example is relatively small and so larger 
Areas of Interest (AOI) would benefit from greater computational power.

Let's get into it...

## Project set up

First let's load the libraries we need for the analysis. If you don't have these
installed, you can run `install.packages(<PACKAGE_NAME>)` to install them. 

```{r}
#| label: load-lib
#| message: false

#Data download and wrangling
library(curl)
library(dplyr)
library(tidyr)

#Machine Learning
library(mlr3verse)
library(mlr3spatiotempcv)
library(mlr3extralearners)
library(mlr3mbo)
library(lightgbm)

#visualisation
library(ggplot2)
library(ggtext)
library(ggpmisc)
library(tmap)
library(gt)
library(scales)

#spatial
library(terra) 
library(sf)
library(rstac)
library(gdalcubes)

```

Now let's set a random seed number - this is important as we're going to  be 
using algorithms (actually just one in this example) that depend on random number 
generation; by setting this value we ensure that the results are reproducible

```{r}
#| label: set-seed
set.seed(5446) 
```

## Training Data 
Now let's download some Tree Canopy Height (TCH) data generated as part of a biomass
modelling paper by @asner2018mapped.  We're going to use this as our
"truth". In reality, this is actually another model, created using Landsat data
and Aerial LiDAR Survey (ALS) data from 2016.
Therefore, this example is more like a model emulation rather than estimating 
actual measured canopy height; but hopefully a useful example - this could be as
easily replaced with any other TCH data. It is quite a large file so might take
a little while...

```{r}
#| label: download-training-data

if (!dir.exists("data")) dir.create("data")

sabah_tch_path <- "data/GAO_TCH_30m_unmasked.tif"
#https://zenodo.org/record/4549461
# file is ~ 800MB so takes a little time.
if (!file.exists(sabah_tch_path)){
  tch.url <- "https://zenodo.org/record/4549461/files/GAO_TCH_30m_unmasked.tif?download=1"
  curl::curl_download(tch.url, destfile = sabah_tch_path, quiet=FALSE)
}

```


Let's focus in on a particular area... I've picked this region of Sabah at random
as it looked like it had a nice diversity of canopy heights and landuse. 
So  Let's define a centre point of our Area of Interest (AOI) and buffer it by 
10km - again this is arbitrary - and solely for this example. Then we crop the 
training TCH data that we downloaded in the last step to be coincident with our AOI.

```{r}
#| label: set-proj-extent
proj.area <- sf::st_point(c(479911, 657243))|> #  # sabah example 
  st_sfc(crs=32650) |> # for sabah 
  st_buffer(10000, endCapStyle = "SQUARE")


tch.proj <- terra::rast(sabah_tch_path) |> 
  terra::crop(proj.area)
# tch.proj[is.na(tch.proj)] <- 0
```


It's always a good idea to visualise what you're modelling - let's take a look...
Here we're plotting the TCH data with the [{tmap}](https://r-tmap.github.io/tmap/) package which uses a Javascript
library called leaflet (in the background) to create an interactive map. FYI, 
if you find this type of interactive mapping useful you should also check out the [{mapview}](https://r-spatial.github.io/mapview/) package which does much the same but
trades of some of the flexibility of tmap for simpler syntax.

```{r}
#| label: interactive-map
#| message: false

#' Generate a base tmap object to display some basemaps.
#'
#' @return A tmap object
tm_basic <- function() {
  tmap::tm_basemap(tmap::providers$Esri.WorldImagery) +
    tmap::tm_basemap(tmap::providers$OpenStreetMap.HOT) +
    tmap::tm_basemap(tmap::providers$Stamen.Terrain) +
    tmap::tm_basemap(tmap::providers$CartoDB.DarkMatter)
}

tmap::tmap_mode("view")
tm_basic() +
  tmap::tm_shape(tch.proj, name = "Tree Canopy Height", raster.downsample = TRUE) +
  tmap::tm_raster(
    palette ="viridis",
    style = "cont",
    title = "Tree Canopy Height"
  )

```

## Earth Observation Covariate Data

Now, there are many ways to explore and download Earth Observation (EO) data.
Many of us use [Google Earth Engine (GEE)](https://earthengine.google.com/) as a 
powerful cloud based platform for data processing. But it can be challenging to 
work both locally and in GEE (although made much easier in R with the [{rgee}](https://r-spatial.github.io/rgee/)). 
Further, not all of the processing we need can be achieved in GEE (although many things can
and there are frequently alternatives). This is as much about showing that there are
alternative and fully open source options for processing/working with EO data.  

So, introducing [Spatio Temporal Asset Catalogs (STAC)](https://stacspec.org/en). 
STAC is basically a way that data providers can store/index their geospatial data 
so that it can be easily searched, filtered and downloaded using a common 
Application Programming Language (API).

### Landsat Composite

In this next code chunk we search the Microsoft Planetary Computer (MPC) STAC 
catalog for Landsat 8 imagery in 2016 for our AOI using the [{rstac} package](https://brazil-data-cube.github.io/rstac/). We don't have to use MPC, 
there are many other catalogs that host the same data but I have found MPC to 
be very performant.

```{r}
#| label: stac-search

# Conver our AOI to Lat Long bounding box
bbox <- st_bbox(tch.proj) 
bbox_wgs84 <- bbox |>
  st_as_sfc(crs=32650) |> 
  st_transform("EPSG:4326") |>
  st_bbox()

# submit get request to STAC endpoint
s = stac("https://planetarycomputer.microsoft.com/api/stac/v1/")
col.list <- rstac::collections(s) |> 
  get_request()


# post the request with initial filtering parameters
items = s |>
  stac_search(collections = "landsat-c2-l2",
              bbox = c(bbox_wgs84["xmin"],bbox_wgs84["ymin"],
                       bbox_wgs84["xmax"],bbox_wgs84["ymax"]), 
              datetime ="2016-01-01T00:00:00Z/2016-12-30T00:00:00Z" )|>
  post_request()|> items_sign(sign_fn = sign_planetary_computer())


length(items$features) # How many images do we have
```
Note how there are `r length(items$features)` images returned in the initial 
search; This includes different sensors with varying cloud coverages. Let's 
refine this firstly by stating the assests (bands) we want to download, 
specify the sensor we want (Landsat 8) and finally the percent of cloud cover 
we will accept as a maximum. 


```{r}
#| label: stac-search-filter
#| warning: false
l8_collection = stac_image_collection(
  items$features,
  asset_names = c(
    "qa_pixel",
    "coastal" ,
    "blue" ,
    "green",
    "red" ,
    "nir08",
    "swir16",
    "swir22"
  )    ,
  property_filter = function(x) {
    # 
     x[["eo:cloud_cover"]] < 20 & x[["platform"]]=="landsat-8"
    #
    #
  }
)
l8_collection
```
Now we have reduced our cube to just `r nrow(gdalcubes:::gc_image_collection_info(l8_collection)$images)` 
images, it's time to create a composite. Here we use the {gdalcubes} package to 
create this, first we define the structure of the cube, i.e. the spatio temporal 
dimensions, coordinate reference system (CRS), temporal reduction method (here 
we use median) and the spatial resampling method.


```{r}
#| label: define-data-cube
e <- extent(l8_collection)

gdalcubes_options(parallel = parallel::detectCores()-4) # set the number of cores to use.
v = cube_view(srs=paste0("EPSG:",st_crs(proj.area)$epsg), dx=30, dy=30, dt="P1Y", 
              aggregation="median", resampling = "bilinear",
              extent=
                list(t0 = e$t0, t1 = e$t1,
                          left=bbox["xmin"], right=bbox["xmax"],
                          top=bbox["ymax"], bottom=bbox["ymin"])
              )
v


```


Now let's download the data and apply cloud masking and some pixel maths in the 
process to generate an annual composite for 2016 with an additional band showing
the Normalised Difference Vegetation Index (NDVI). 

```{r}
#| label: download-EO-data

# L8.clear_mask = image_mask("qa_pixel", bits=1:4, values=16)
L8.clear_mask = image_mask("qa_pixel", values=21824, invert=TRUE)
cube <- raster_cube(l8_collection, v, mask = L8.clear_mask) |> 
  apply_pixel(c("(nir08-red)/(nir08+red)"), names="NDVI", keep_bands=TRUE) 

if(!dir.exists("out_data")) dir.create("out_data")

terra_cube <-
  write_tif(cube, dir = "data_out", prefix = "Sabah_example_") |>  #
  rast()

# Rescale the imagery but not NDVI that one is fine.
terra_cube[[setdiff(names(terra_cube), c("NDVI", "qa_pixel"))]] <- 
  terra_cube[[setdiff(names(terra_cube), c("NDVI", "qa_pixel"))]]/10000 


```


What did we just download? Let's take a look... You'll see that to generate to 
RGB image in R, a little extra work is needed to get a visually recognisable 
image. You can use `terra::plotRGB` but the colours might look a bit "off". 
`tmap::tm_rgb` can also be used but ggplot does offer a lot of control so it is 
useful to know.

```{r}
#| label: ggplot-rgb

#' Convert an RGB raster to a dataframe ready to plot with ggplot.
#'
#' @param r SpatRaster object that contains bands named red, green and blue.
#' @param .min Numeric between 0 and 1. The minimum cutoff value to visualise.
#' @param .max Numeric between 0 and 1. The maximum cutoff value to visualise.
#' @param .yr Adds a labelling column - in this case the year.
#'
#' @return A dataframe representing the values of the rgb raster with x and y locations.
l8_to_rgb_df <- function(r, .min=0.001, .max=1.2, .yr=2016){
  as.data.frame(r, xy=TRUE) |>
    mutate(red = case_when(red<.min ~ .min,
                          red>.max ~ .max,
                          TRUE ~ red),
           red =scales::rescale(red, c(0,1)),
           green = case_when(green<.min ~ .min,
                          green>.max ~ .max,
                          TRUE ~ green),
           green = scales::rescale(green, c(0,1)),
           blue = case_when(blue<.min ~ .min,
                          blue>.max ~ .max,
                          TRUE ~ blue),
           blue = scales::rescale(blue, c(0,1)),
           Year = .yr)

}

#' create an RGB raster map with {ggplot2}
#'
#' @param .x SpatRaster object that contains bands named red, green and blue.
#'
#' @return A ggplot
rgb_plot <- function(.x){
  
  l8_df <- l8_to_rgb_df(.x[[c("red", "green", "blue")]]) |>  # run the function to get a dataframe
    tidyr::drop_na()
            
  ggplot(data=l8_df, aes(x=x, y=y, fill=rgb(red,green,blue))) +
  geom_raster() +
  scale_fill_identity() +
theme_light() +
  scale_x_continuous(breaks = round(seq(min(l8_df$x), max(l8_df$x), by = 5e+3),0)) +
  scale_y_continuous(breaks = round(seq(min(l8_df$y), max(l8_df$y), by = 5e+3),0))+
  theme(
    axis.title.x=element_blank(),
    axis.title.y=element_blank()) +
  coord_fixed()
}

rgb_plot(terra_cube)

```

### Terrain Data

In addition to the spectral data from our Landsat 8 composite, we might also 
want to include other environmental covariates which could have some relationship
with tree canopy height. So, let's add some terrain data and related geomorphometrics.
We again use `rstac` to retrieve the DTM tiles. Here we don't need to use `gdalcubes`
because the global DTM is assumed to represent a snapshot in time and therefore we
do not have overlapping tiles. Here, `terra` (which uses [gdal](https://gdal.org/)
for reading data) is very capable of reading on-line GeoTiff files from urls.


```{r}
#| label: get-terrain-data

#' Get the URLs for DTM tiles.
#'
#' @param aoi_box A bbox object (using EPDG:4326 CRS)
#' @param collection default "cop-dem-glo-30". The DTM collection to request.
#'
#' @return A list of DTM tile urls from the Microsoft Planetary Computer STAC catalog.
mpc_dtm_src <- function(aoi_box,
                        collection = "cop-dem-glo-30"){
  
  s_obj <- rstac::stac("https://planetarycomputer.microsoft.com/api/stac/v1")
  rstac::get_request(s_obj)
  
  it_obj <- s_obj  |>
    rstac::stac_search(collections = collection[1],
                       bbox = c(aoi_box["xmin"],aoi_box["ymin"],
                                aoi_box["xmax"],aoi_box["ymax"]))  |>
    rstac::get_request()
  
  src_list <-rstac::assets_url(it_obj)
  
  .urls <- src_list[grep(".tif$", src_list)]
  
  sapply(.urls, function(x) paste0("/vsicurl/", x), USE.NAMES =FALSE)
  
}

tch.reproj <- project(tch.proj, terra_cube)

dtm <- mpc_dtm_src(bbox_wgs84) |> # call the function to get the tile urls
  lapply(rast) |> # iterate over the tiles and load as a spatRaster object
  terra::sprc() |> # convert to a collection
  terra::merge() |> # merge the tiles - basic mosaic - nothing overlaps here so this is fine.
  project(tch.reproj) # project into our desired CRS and extent etc.
names(dtm) <- "dtm"
# now generate a few geomorphometric layers
asp <- terra::terrain(dtm, "aspect")
slp <- terrain(dtm, "slope")
TRI <- terrain(dtm, "TRIrmsd")
rough <- terrain(dtm, "roughness")
```

## Building The Model Inputs

Now, let's combine the Terrain and Landsat 8 composite data into a single raster with all 
of the bands. Then we convert this to a dataframe so that we can apply the 
statistical models as we need.

```{r}
#| label: build-cube

# combine all of our layers from L8 and the terrain.
tch.reproj <- project(tch.proj, terra_cube)

comb_dat <- c(tch.reproj,terra_cube, dtm, asp, slp, TRI, rough) 
drop_bands <- c("qa_pixel") # we don't want to model with pixel masks so drop it. Also dropping lwir11 as there are a lot of missing values and not useful for our needs.
comb_dat <- comb_dat[[setdiff(names(comb_dat), drop_bands)]]
names(comb_dat[[1]]) <- "TCH"
# this will be our dataframe for the ML section
comb_df <- as.data.frame(comb_dat, xy=TRUE) |> 
  tidyr::drop_na()




```


Before we start modelling, what did we just download? Let's take a look...Here 
are the individual bands, plotted using `ggplot2` and `patchwork`. Again,
a simple `plot(comb_dat)`will work OK but does not correctly display the data 
due to the differences in ranges across bands.

```{r}
# plotting data in ggplot requires longform data.
comb_df |> 
  tidyr::pivot_longer(-one_of(c("x", "y"))) |> # converts df to longform.
  dplyr::group_by(name) |> # groups by band
  dplyr::group_map(~ggplot(.x) + # applies plot function by band
  aes(x=x, y=y, fill=value)+
  geom_raster() +
    guides(fill="none") +
  scale_fill_viridis_c() +
    coord_fixed()+
    theme_void() +
    labs(subtitle=.x$name[1]) +
  theme(
    axis.title.x=element_blank(),
    axis.title.y=element_blank(),
    axis.text.x=element_blank(),
    axis.text.y=element_blank()),
  .keep=TRUE) |> 
    patchwork::wrap_plots(nrow=3) # combines all plots into a grid

```
## Machine Learning with {mlr3}

Now let's do some statistics/machine learning - so much jargon in this space!
we're going to use the [{mlr3}](https://mlr3.mlr-org.com/) ecosystem to simplify 
our workflow - it has lots of functionality for applying different ML techniques 
and algorithms. This example is very basic but check out the [mlr3 book](https://mlr3book.mlr-org.com/)
for loads of great infotmation. You may also want to consider the [tidymodels](https://www.tidymodels.org/)
ecosystem which has very similar functionality to mlr3 but is stylistically 
very different. 

The first step with `mlr3` is to define the `task` this is essentially a template/
instruction set for the other mlr3 functions to work with. 

```{r}
#  ---------- GENERATE A "TASK" ------------
task = mlr3spatiotempcv::TaskRegrST$new(
  id = "Sabah_CHM",
  backend = comb_df, 
  target = "TCH", 
  coordinate_names = c("x", "y"),
  extra_args = list(
    coords_as_features = FALSE,
    crs = terra::crs(comb_dat))
)
```

You'll note that this task is a little bit special because we use the [mlr3spatiotempcv package](https://mlr3spatiotempcv.mlr-org.com/articles/mlr3spatiotempcv.html);
essentially this just manages the additional spatial/temporal attributes of our data: 
coordinates, CRS and time (if we were doing multi-temporal work which we aren't here)
to manage resamapling strategies.

So, let's define the spatial resampling method that we will use to evaluate 
our model's performance. We can also use this for model tuning (although this
is not included in this example). Note how training and test data partitions 
are spatially distinct. There are many different ways to do this spatial partitioning
but this method is simple and often more than suitable. It uses kmeans 
clustering to split the AOI into six regions (based on the requested 
number of folds). It is important to do this to avoid over estimating our model's
performance by evaluating predictions with neighbouring or nearby cells which
are more likely to be similar due to their spatial location. 

```{r}

# ---------- DEFINTE A RESAMPLING PLAN ------------
spcv_plan <- mlr3::rsmp("repeated_spcv_coords", folds = 6, repeats=1)

autoplot(spcv_plan, task = task, fold_id=1:3)

```

Now we set some "learners" (AKA algorithms) - let's test two in this case LightGBM
and a Generalised Linear Model (GLM). 

```{r}
# ------- DEFINE SOME LEARNERS
lgbm.lrn <- mlr3::lrn("regr.lightgbm")
lrn.glm <- mlr3::lrn("regr.glm")
```


Let's benchmark these learners and see which will perform better:

```{r}
design = benchmark_grid(task, list(lgbm.lrn, lrn.glm), spcv_plan)
# print(design)
future::plan("multisession", workers = 6) # sets the number of cores to run on -  we have
bmr = mlr3::benchmark(design)
aggr = bmr$aggregate(measures = c(msr("regr.rmse"), msr("regr.mse"), msr("regr.rsq")))


gt::gt(aggr[,4:9,])

```

So the Lightgbm model looks to be the best - we could very well get a better performance out of
all these models with some tuning and feature selection but, for now, let's keep
it simple. The result from the LightGBM also looks pretty good.

So, let's check this in more detail by running the resampling and inspecting the results
by plotting the predicted sets from the resampling:


```{r}
resample_lgbm <- progressr::with_progress(expr = {
  mlr3::resample(
    task = task,
    learner = lgbm.lrn, 
    resampling = spcv_plan,
    store_models = FALSE,
    encapsulate = "evaluate"
  )
})

resample_lgbm$aggregate(measure = c(
    mlr3::msr("regr.bias"),
    mlr3::msr("regr.rmse"),
    mlr3::msr("regr.mse")))

resample_lgbm$prediction() |> 
  ggplot() +
  aes(x=response, y=truth)+
  geom_bin_2d(binwidth = 0.3) +
  scale_fill_viridis_c(trans=scales::yj_trans(0.1), option="G", direction=-1) +
  geom_abline(slope=1) +
  theme_light()

```

Great, so we have our chosen model, let's predict across our full cube using all available
data - we've done our accuracy evaluation so it makes no sense to have any data held out
for the final model. Training and predicting the model is just two lines of code:

```{r}
#| label: train-and-predict

#---------- TRAIN THE MODEL -----------
lgbm.lrn$train(task)


#----------- PREDICT THE FULL MODEL --------
p <- terra::predict(comb_dat, lgbm.lrn, na.rm=FALSE)

```

Now, remember that the Tree canopy height data we are emulating is a model itself,
also, we are not totally sure what imagery was used to generate it. So, given this,
take a look at the differences between the original TCH model and our LightGBM model results and 
consider why there are some fairly curious differences...

```{r}
#| label: plot-and-compare-model
#| warning: false
#----------- Let's compare with the training data --------

l8rgb <- terra_cube[[c("red", "green", "blue")]]

l8rgb.mat <- as.matrix(l8rgb) |>
  scales::rescale()
values(l8rgb) <- l8rgb.mat
tm_basic() +
  tm_shape(l8rgb, name="L8-rgb", raster.downsample = FALSE) +
  tm_rgb(r= 1, g=2, b=3, max.value = 1)+
  tmap::tm_shape(tch.reproj, name = "Tree Canopy Height", raster.downsample = TRUE) +
  tmap::tm_raster(
    palette ="viridis",
    style = "cont",
    title = "Tree Canopy Height",
    breaks=c(seq(0, 50, 10))
  ) +
  tmap::tm_shape(p, name = "Modelled Tree Canopy Height", raster.downsample = FALSE) +
  tmap::tm_raster(
    palette ="viridis",
    style = "cont",
    title = "Modelled Tree Canopy Height",
    breaks=c(seq(0, 50, 10))
  )
```


## Summary

- We have downloaded some training data to use as an input to train
our own machine learning model. 

- Then we downloaded Landsat 8 data and created a cloud masked annual median 
composite for 2016. 

- Next, we downloaded a dtm (Copernicus 30 m) and derived a range of geomorphometric 
covariates. 

- We combined all of these bands of data into a single raster,  and then
created a data frame comprising all of these values (along with our training TCH data).

- Then, we tested two ML models using their default values, benchmarked their
predictive performance and, based on this benchmark, selected the LightGBM model
as having better performance.

- We generated resampled predictions using spatially distinct test/train sets and 
visualised these results alongside generating aggregated model evaluation metrics.

- Finally, we trained the model and predicted these values across our data cube to 
create our own TCH model. In prinicple we could now apply this model into new years
and new regions, taking care to ensure that the new regions are reasonably described by
the environmental covariates in this training process.


## Further considerations

We have done no model tuning or feature selection. Both of these things can be
carried out in mlr3 and are very likely to yield significant model improvements
but will take a little longer to run. 

Other ML algorithms may also yield good or even better results than presented
here However, LightGBM is very fast and does often generate good results for these
sorts ML projects. 

We haven't applied our model into novel regions or time periods - if we chose to
do this, there are some important things to consider such as ["area of applicability"](https://hannameyer.github.io/CAST/articles/cast02-AOA-tutorial.html#:~:text=The%20AOA%20is%20defined%20as,estimated%20cross%2Dvalidation%20performance%20holds.) and 
spectral consistency across annual composites. we also haven't included any 
data that might more directly describe vegetations structure in this example,
data such as [Synthetic Aperture Radar (SAR)](https://www.earthdata.nasa.gov/learn/backgrounders/what-is-sar) 
might help to improve model performance but SAR comes with many unique challenges. 








